Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-26 Manuscript/manuscript.Rmd
edits

We then shifted our focus toward developing a method to more accurately identify the location of TAD boundaries throughout the linear genome. Typical TAD callers are restricted to kilobase resolutions, which increase the variability in the actual location of a boundary. Instead, we aimed to predict the location of TAD boundaries at basepair resolution using optimally built cell line specific random forest classifiers. Each model was restricted to only include the most influential cell line specific TFBS, namely CTCF, RAD21, SMC3a, adn ZNF143. We then validated our models on cell line and chromosome specific regions of the genome, at base pair resolution. Predicted TAD boundary regions (PTBR) were identified as the collection of base pairs where the predicted probability vector was equal to 1, which denoted 100% accuracy in correctly identifying a basepair as a potential TAD boundary (Figure 7A). To allow for minor fluctuations in probability, any base pair given probability of 1 that was within a 10 bp window from another base pair also given probability of 1 were combined into one PTBR. PTBR were then compared to the called TAD boundary regions (CTBR), which here were defined by a 10 kb resolution flanking (See Methods). There were a total of 4663 identified PTBR compared to 237 CTBR. We found that approximately 59% (139 out of 237) of CTBR were overlapped by PTBR, indicating there are fluctuations in probabilities within the 20 kb regions defining CTBR (Figure 7B). We then compared distances between PTBRs and annotations versus CRBRs and annotations. We found that boundary regions predicted by our model at bp resolution experienced short distance profiles to known architectural proteins than did boundary regions called by Arrowhead.

![**Comparing TAD boundaries regions predicted at bp resolution with those called at 10 kb resolution.** (A) A subset of the probabilitity vector for predicting TAD boundaries for chr22:17380000-chr22:18650000. The red dashed lines represent predicted TAD boundary regions, while the blue dashed lines represent called TAD boundaries identified by Arrowead at 10 kb. A called TAD boundary region is one flanked by 10 kb (B) Venn diagram illustrating the number of overlapped regions between predicted (red) and called (blue) TAD boundary regions. (C) Boxplots illustrating the distance profiles between each class of boundary regions (predicted-red; called-blue) and architechtural proteins known to be enriched at TAD boundaries.](Figures/pred_at_bp_res_plots.png)

>>>>

After obtaining our optimally built random forest classifier, we aimed to leverage the ChIP-seq peak regions defined by the SMC3, ZNF143, CTCF, and RAD21 tanscription factors for precised TAD bounday prediction at base pair resolution. We found that when we validated our model against chromosome specific base pairs, we were able to identify clusters of regions where the model predicted the base pair to be a TAD boundary with probability equal to 1. By clustering the regions and performing medoid partitioning (See Methods), we were able to identify a comparative number of TAD boundaries as the ARROWHEAD TAD caller. However, only 24% and 29% of predicted and called (and 10 kb flanked) TAD boundaries overlapped for the GM12878 and K562 cell line respectively, suggesting differing locations in TAD boundaries throughout the linear genome (Supplementary Figure 10). To compare between predicted and called TAD boundaries we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flank of each. We found that the peak signal strength for the SMC3, ZNF143, CTCF, and RAD21 TFBS were much closer, on average, to our predicted TAD boudaries the the called TAD boundaries obtained using the ARROWHEAD pipeline for 10 kb resolution Hi-C data (Figure 5; Supplementary Figure ). ChIP-seq assays offer much higher resolutions (often in the tens of bases) compared to Hi-C sequenced data, which can be in the hundreds of kilobases. This disparity in resolution can affect a TAD caller's ability to precisely locate TADs. Moreover, by building a model using annotations known to be enriched at TAD boundaries we were able to bridge this gap in resolution and more precisely locate TAD boundaries that corresponded to the highest concentration in signal strength for top ranked TFBS.

![**Predicted TAD boundaries were centered closer to peak signal regions defined by known architectural transcription factors when compared to called TAD boundaries.** Enrichment heatmaps illustrate the average signal strength of ChIP-seq peak regions defined by (A) SMC3, (B) ZNF143, (C) CTCF, and (D) RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.](Figures/enrichment_heatmaps_gm12878.png)
